---
title: "Stat 151a final project"
author: "Martin Garcia-Angel"
date: "2023-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(magrittr)
library(ggplot2)
library(faux)
```

```{r}
hour =read.csv("C:\\Users\\Marti\\Downloads\\hour.csv")
head(hour)
```

```{r}
day =read.csv("C:\\Users\\Marti\\Downloads\\day.csv")
head(day)
```

```{r}
nrow(hour)
nrow(day)
```

### Line plot showing bike usage per day

```{r}
# Create the line plot
ggplot(day, aes(x = instant, y = cnt)) +
  geom_line() +
  labs(x = "Index", y = "Count bike usage per Day", title = "Total Bike usage over Time") +
  theme_minimal()
```

### Checking total count of bikes per season

```{r}
ggplot(day, aes(x = season, y = cnt)) + geom_bar(stat = 'identity')+ scale_y_continuous(labels = scales::comma)
```

## Checking total number of bikes per every month of the year

```{r}
day$mnth <- as.factor(day$mnth)

ggplot(day, aes(x = factor(mnth), y = cnt)) + 
  geom_bar(stat = 'identity') +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Month", y = "Count")
```

### box plot of Daily usage during Holiday vs Non-Holiday

```{r}
box_plot <- ggplot(day, aes(x = factor(holiday), y = cnt, fill = factor(holiday))) +
  geom_boxplot() +
  labs(x = "Holiday", y = "Count", title = "Daily Usage for Holiday vs Non-Holiday") +
  scale_fill_manual(values = c("pink", "skyblue"))  # Specify colors for boxes
box_plot
```

### box plot of Hourly usage during Holiday vs Non-Holiday

```{r}
box_plot <- ggplot(hour, aes(x = factor(holiday), y = cnt, fill = factor(holiday))) +
  geom_boxplot() +
  labs(x = "Holiday", y = "Count", title = "Hour Usage for Holiday vs Non-Holiday") +
  scale_fill_manual(values = c("pink", "skyblue"))  # Specify colors for boxes
box_plot
```

### Bar plot of average ridership for holiday vs non-holiday

```{r}
avg_counts <- day %>%
  group_by(holiday) %>%
  summarise(avg_count = mean(cnt, na.rm = TRUE))

bar_chart <- ggplot(avg_counts, aes(x = factor(holiday), y = avg_count, fill = factor(holiday))) +
  geom_bar(stat = 'identity') +
  labs(x = "Holiday", y = "Average Count", title = "Average Daily Count of Bike Usage for Holiday vs Non-Holiday") +
  scale_fill_manual(values = c("pink", "skyblue"))

bar_chart
```

```{r}
# Calculate average counts per hour
avg_counts <- day %>%
  group_by(season) %>%
  summarise(avg_count = mean(cnt, na.rm = TRUE))

# Create the line plot
ggplot(avg_counts, aes(x = season, y = avg_count)) +
  geom_point() +
  geom_line() +
  labs(x = "Season", y = "Average Count per day in the Season", title = "Average Count of Everyday Bike Usage per season") +
  theme_minimal()
```


### Bar plot of average ridership for per day of the week

```{r}
avg_counts <- day %>%
  group_by(weekday) %>%
  summarise(avg_count = mean(cnt, na.rm = TRUE))

# Reorder the levels of 'weekday' based on average count (descending order)
avg_counts <- avg_counts %>%
  mutate(weekday = reorder(factor(weekday), -avg_count))

# Create the bar chart
ggplot(avg_counts, aes(x = weekday, y = avg_count, fill = factor(weekday))) +
  geom_bar(stat = 'identity') +
  labs(x = "Day of the Week", y = "Average Count", title = "Average Count Comparison: Day of the Week") 

```

### Box plot of Ridership for per day of the week

```{r}
ggplot(day, aes(x = factor(weekday), y = cnt, fill = factor(weekday))) +
  geom_boxplot() +
  labs(x = "Weekday", y = "Count", title = "Count Distribution: Day of the Week")
```

### Box plot of Ridership for per Workingday

```{r}
ggplot(day, aes(x = factor(workingday), y = cnt, fill = factor(workingday))) +
  geom_boxplot() +
  labs(x = "Workingday", y = "Count", title = "Count Distribution: Working Day?")
```

## Total number of riders during 2011 and 2012

```{r}
ggplot(day, aes(x = factor(yr), y = cnt, fill = factor(yr))) +
  geom_bar(stat = 'identity') +
  labs(x = "Year", y = "Count", title = "Count Distribution: Year") +
  scale_fill_manual(values = c("pink", "skyblue"))  # Specify colors for boxes
```

### Registered Riders vs Total count

```{r}
ggplot(day, aes( x = registered, y = cnt)) + geom_point() + labs(x = 'Registered Riders', y = 'Total Count Riders', title = 'Registered Riders vs Total Count') 

```

### Pairs plot of all numerical variables

```{r}
day_data_2 = day[, c('cnt','registered', 'casual', 'temp', 'hum', 'windspeed')]
pairs(day_data_2)
```

### Average bike usage per hour of the day

```{r}
# Calculate average counts per hour
avg_counts <- hour %>%
  group_by(hr) %>%
  summarise(avg_count = mean(cnt, na.rm = TRUE))

# Create the line plot
ggplot(avg_counts, aes(x = hr, y = avg_count)) +
  geom_point() +
  geom_line() +
  labs(x = "Hour of the Day", y = "Average Count of Bike Usage", title = "Average Count of Bike Usage per Hour") +
  theme_minimal()
```

### Median bike usage per month of the Year

```{r}
# Calculate average counts per hour
median_count <- hour %>%
  group_by(mnth) %>%
  summarise(avg_count = median(cnt, na.rm = TRUE))

# Create the line plot
ggplot(median_count, aes(x = mnth, y = avg_count)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = 1:12, labels = 1:12) +
  labs(x = "Month of the Year", y = "Median Count of Bike Usage", title = "Median Count of Bike Usage for Each month of the Year") +
  theme_minimal()
```

## ONE HOT ENCODING

```{r}

# Create binary columns for each season
data <- hour

# Create binary columns for each season
data <- data %>%
  mutate(
    Season1 = ifelse(season == 1, 1, 0),
    Season2 = ifelse(season == 2, 1, 0),
    Season3 = ifelse(season == 3, 1, 0),
    Season4 = ifelse(season == 4, 1, 0)
  )

# Create binary columns for each season
data <- data %>%
  mutate(
    Weather1 = ifelse(weathersit == 1, 1, 0),
    Weather2 = ifelse(weathersit == 2, 1, 0),
    Weather3 = ifelse(weathersit == 3, 1, 0),
    Weather4 = ifelse(weathersit == 4, 1, 0)
  )

data <- data %>%
  mutate(
    Weekday0 = ifelse(weekday == 0, 1, 0),
    Weekday1 = ifelse(weekday == 1, 1, 0),
    Weekday2 = ifelse(weekday == 2, 1, 0),
    Weekday3 = ifelse(weekday == 3, 1, 0),
    Weekday4 = ifelse(weekday == 4, 1, 0),
    Weekday5 = ifelse(weekday == 5, 1, 0),
    Weekday6 = ifelse(weekday == 6, 1, 0)
  )
data <- data %>%
  mutate(
    hr0 = ifelse(hr == 0, 1, 0),
    hr1 = ifelse(hr == 1, 1, 0),
    hr2 = ifelse(hr == 2, 1, 0),
    hr3 = ifelse(hr == 3, 1, 0),
    hr4 = ifelse(hr == 4, 1, 0),
    hr5 = ifelse(hr == 5, 1, 0),
    hr6 = ifelse(hr == 6, 1, 0),
    hr7 = ifelse(hr == 7, 1, 0),
    hr8 = ifelse(hr == 8, 1, 0),
    hr9 = ifelse(hr == 9, 1, 0),
    hr10 = ifelse(hr == 10, 1, 0),
    hr11= ifelse(hr == 11, 1, 0),
    hr12 = ifelse(hr == 12, 1, 0),
    hr13 = ifelse(hr == 13, 1, 0),
    hr14 = ifelse(hr == 14, 1, 0),
    hr15 = ifelse(hr == 15, 1, 0),
    hr16 = ifelse(hr == 16, 1, 0),
    hr17 = ifelse(hr == 17, 1, 0),
    hr18 = ifelse(hr == 18, 1, 0),
    hr19 = ifelse(hr == 19, 1, 0),
    hr20 = ifelse(hr == 20, 1, 0),
    hr21 = ifelse(hr == 21, 1, 0),
    hr22 = ifelse(hr == 22, 1, 0),
    hr23 = ifelse(hr == 23, 1, 0)
  )
data <- data %>%
  mutate(
    month1 = ifelse(mnth == 1, 1, 0),
    month2 = ifelse(mnth == 2, 1, 0),
    month3 = ifelse(mnth == 3, 1, 0),
    month4 = ifelse(mnth == 4, 1, 0),
    month5 = ifelse(mnth == 5, 1, 0),
    month6 = ifelse(mnth == 6, 1, 0),
    month7 = ifelse(mnth == 7, 1, 0),
    month8 = ifelse(mnth == 8, 1, 0),
    month9 = ifelse(mnth == 9, 1, 0),
    month10 = ifelse(mnth == 10, 1, 0),
    month11 = ifelse(mnth == 11, 1, 0),
    month12 = ifelse(mnth == 12, 1, 0)
  )
```

##OHE for DAY
```{r}
# Create binary columns for each season
data1<- day

# Create binary columns for each season
data1 <- data1 %>%
  mutate(
    Season1 = ifelse(season == 1, 1, 0),
    Season2 = ifelse(season == 2, 1, 0),
    Season3 = ifelse(season == 3, 1, 0),
    Season4 = ifelse(season == 4, 1, 0)
  )

# Create binary columns for each season
data1 <- data1 %>%
  mutate(
    Weather1 = ifelse(weathersit == 1, 1, 0),
    Weather2 = ifelse(weathersit == 2, 1, 0),
    Weather3 = ifelse(weathersit == 3, 1, 0),
    Weather4 = ifelse(weathersit == 4, 1, 0)
  )

data1 <- data1 %>%
  mutate(
    Weekday0 = ifelse(weekday == 0, 1, 0),
    Weekday1 = ifelse(weekday == 1, 1, 0),
    Weekday2 = ifelse(weekday == 2, 1, 0),
    Weekday3 = ifelse(weekday == 3, 1, 0),
    Weekday4 = ifelse(weekday == 4, 1, 0),
    Weekday5 = ifelse(weekday == 5, 1, 0),
    Weekday6 = ifelse(weekday == 6, 1, 0)
  )

data1 <- data1 %>%
  mutate(
    month1 = ifelse(mnth == 1, 1, 0),
    month2 = ifelse(mnth == 2, 1, 0),
    month3 = ifelse(mnth == 3, 1, 0),
    month4 = ifelse(mnth == 4, 1, 0),
    month5 = ifelse(mnth == 5, 1, 0),
    month6 = ifelse(mnth == 6, 1, 0),
    month7 = ifelse(mnth == 7, 1, 0),
    month8 = ifelse(mnth == 8, 1, 0),
    month9 = ifelse(mnth == 9, 1, 0),
    month10 = ifelse(mnth == 10, 1, 0),
    month11 = ifelse(mnth == 11, 1, 0),
    month12 = ifelse(mnth == 12, 1, 0)
  )
```

-Testing the best subset for our model on all categorical variables

```{r}
names(data)
library(leaps)
data$hrssn= data$season * data$workingday 
data$weathersitwork= data$hr8 * data$weathersit 
bhour <- regsubsets(cnt~registered+Season1+Season2+Season3+Season4+Weekday0+Weekday1+Weekday2+Weekday3+Weekday4+Weekday5+Weekday6+Weather1+Weather2+Weather3+Weather4
                    +hr1+hr2+hr3+hr4+hr5+hr6+hr7+hr8+hr9+hr10+hr11+hr12+hr13+hr14+hr15+hr16+hr17+hr18+hr19+hr20+hr21+hr22+hr23+month1+month2
                    +month3+month4+month5+month6+month7+month8+month9+month10+month11+month12+holiday+workingday+yr+temp+windspeed+atemp+hum+weathersitwork+hrssn, data,really.big=T)
summary(bhour)
summary(bhour)$adjr2
which.max(summary(bhour)$adjr2)
```



```{r}
day1 = subset(day, yr == 1)
hour$hrssn= hour$season * hour$workingday 
hour$hrmnth= hour$workingday * hour$weathersit 
big_model = lm(cnt ~registered+ hrssn +hrmnth+temp +hum+workingday, data = hour)
summary(big_model)
plot(big_model)
```
```{r}
library(caret)
day_model = lm(cnt ~registered+ mnth +weathersit+temp +hum+workingday+season, data = day)
predicted_values <- predict(day_model, newdata = day)
RMSE(predicted_values, day$cnt)

comparison_data <- data.frame(
  Instant = day$instant,
  Observed = day$cnt,
  Predicted = predicted_values
)
filtered_data <- comparison_data %>% filter(row_number() %% 10 == 0)

ggplot(filtered_data, aes(x = Instant, y = Observed)) +
  geom_point() +
  geom_line(aes(x = Instant, y = Predicted)) +
  labs(x = "Index", y = "Count", title = "Observed vs. Predicted Count of Bike Usage") +
  theme_minimal()
```

```{r}
subset(comparison_data, Predicted < 0)
comparison_data
```

##log t
```{r}
day_model_logged = lm(log(cnt) ~log(registered)+ mnth +weathersit + log(temp) +hum+workingday+season, data = day)

plot(day_model_logged)
predicted_values <- predict(day_model_logged, newdata = day)
RMSE(predicted_values, log(day$cnt))

comparison_data <- data.frame(
  Instant = day$instant,
  Observed = log(day$cnt),
  Predicted = predicted_values
)
filtered_data <- comparison_data %>% filter(row_number() %% 10 == 0)

ggplot(filtered_data, aes(x = Instant, y = Observed)) +
  geom_point() +
  geom_line(aes(x = Instant, y = Predicted)) +
  labs(x = "Index", y = "Count", title = "Observed vs. Predicted Count of Bike Usage") +
  theme_minimal()
```

```{r}
selected_years <- c(1, 448, 185, 463, 668, 669, 693)

subset(day, instant %in% selected_years , select = c('dteday', 'workingday','weathersit', 'temp', 'hum', 'windspeed','holiday', 'registered', 'cnt')  )
smallest_cnt <- day[which.min(day$cnt), ]
largest_cnt <- day[which.max(day$cnt), ]
largest_cnt
smallest_cnt
```



##standardize
```{r}
data1$tempstd = (day$temp - mean(day$temp))/sd(day$temp)
data1$humstd = (day$hum - mean(day$hum))/sd(day$hum)
data1$registeredstd = (day$registered - mean(day$registered))/sd(day$registered)
data1$cntstd = (day$cnt - mean(day$cnt))/sd(day$cnt)

day_model_standardized = lm(cntstd ~registeredstd+ mnth +weathersit + tempstd +humstd+workingday+ (workingday * tempstd), data = data1)
summary(day_model_standardized)
plot(day_model_standardized)
predicted_values <- predict(day_model_standardized, newdata = data1)
RMSE(predicted_values, data1$cntstd)

comparison_data <- data.frame(
  Instant = day$instant,
  Observed = data1$cntstd,
  Predicted = predicted_values
)
filtered_data <- comparison_data %>% filter(row_number() %% 10 == 0)

ggplot(filtered_data, aes(x = Instant, y = Observed)) +
  geom_point() +
  geom_line(aes(x = Instant, y = Predicted)) +
  labs(x = "Index", y = "Count", title = "Observed vs. Predicted Count of Bike Usage") +
  theme_minimal()

```


```{r}
names(day)
bday <- regsubsets(cntstd~registeredstd+season+Weather1+Weather2+Weather3+Weather4+holiday+yr+month1+month2+month3+month4+month5+month6+month7+month8+month9+month10+month11+month12+Weekday1+Weekday2+Weekday3+Weekday4+Weekday5+Weekday6+Weekday0+tempstd+windspeed+humstd+workingday + Season1+Season2+Season3+Season4+(workingday * tempstd)+(humstd * tempstd), data1)
summary(bday)
summary(bday)$adjr2
which.max(summary(bday)$adjr2)
```



```{r}

big_std_model = lm(cntstd~registeredstd+holiday+month3+month10+Weekday5+tempstd+humstd+workingday +Season2+(workingday * tempstd)+(registeredstd * workingday)+
                     (month3 * tempstd)+(holiday * registeredstd)+(holiday * tempstd)+(registeredstd * Season3), data1)

summary(big_std_model)

plot(big_std_model)
predicted_values <- predict(big_std_model, newdata = data1)
RMSE(predicted_values, data1$cntstd)

comparison_data <- data.frame(
  Instant = data1$instant,
  Observed = data1$cntstd,
  Predicted = predicted_values
)
filtered_data <- comparison_data %>% filter(row_number() %% 10 == 0)

ggplot(filtered_data, aes(x = Instant, y = Observed)) +
  geom_point() +
  geom_line(aes(x = Instant, y = Predicted)) +
  labs(x = "Index", y = "Count", title = "Observed vs. Predicted Count of Bike Usage") +
  theme_minimal()
```
```{r}

day_model_unstd =  lm(cnt~registered+month3+month10+Weekday5+hum+workingday +Season2+(workingday * temp)+(registered * workingday)+
                     (month3 * temp)+(holiday * registered)+(holiday * temp)+(registered * Season3), data1)
summary(day_model_unstd)
plot(day_model_unstd)
predicted_values <- predict(day_model_unstd, newdata = data1)
RMSE(predicted_values, data1$cnt)

comparison_data <- data.frame(
  Instant = data1$instant,
  Observed = data1$cnt,
  Predicted = predicted_values
)
filtered_data <- comparison_data %>% filter(row_number() %% 10 == 0)

ggplot(filtered_data, aes(x = Instant, y = Observed)) +
  geom_point() +
  geom_line(aes(x = Instant, y = Predicted)) +
  labs(x = "Index", y = "Count", title = "Observed vs. Predicted Count of Bike Usage") +
  theme_minimal()
```



